One-armed spirals in locally isothermal, radially structured self-gravitating discs

### 2014-11-26 paper/results.tex
second draft

\subsection{Reference run with random perturbations}
To obtain a picture of the overall disc evolution, we describe a 
fiducial run initialised with $M=10$ in Eq. \ref{randpert}. 
Fig. \ref{fargo_modeamp} plots evolution of the maximum 
non-axisymmetric surface density amplitudes in $R\in[R_{1},R_{2}]$
for $m\in[1,10]$. Snapshots from the simulation are shown in
Fig. \ref{fargo_2d}. 
At early times $t\lesssim100P_0$ the disc 
dominated by low-amplitude high-$m$ perturbations. The $m\geq4$ modes
growth initially and saturate (or decays) after $t=40P_0$. Notice the
low $m\leq 2$ modes decay initially, but grows between $t\in[20,40]P_0$,
possibly due to non-linear interaction of the high-$m$ modes  
\citep{laughlin96,laughlin97}. However, the $m=1$ mode begins to grow
again after $t=70P_0$, and eventually dominates the annulus.

@@

\subsection{Trapping the low-frequency $m=1$ spiral}\label{fargo_m1}
In order to focus on the $m=1$ spiral that eventually emerges, from
now on we examine simulations initialised 
$M=1$ in Eq. \ref{randpert}. The disc 
evolution is largely similar to the previous case, but the one-armed spiral is
more coherent, which facilitates the analysis below.

Fig. \ref{2d_fargo_viz} shows the $m=1$ surface density
at the end of this simulation. For this one-armed spiral
we measure a co-rotation radius and growth rate  
\begin{align*}
  &R_c \simeq 4.4R_0,\\
  &\gamma\simeq 0.014\Omega_k(R_0) = 0.13\Omega_p. 
\end{align*}
The spiral can be considered as low frequency because its 
pattern speed $\Omega_p \simeq 0.1\Omega_k(R_0)\lesssim 0.3\Omega$ in
$R\in[R_{1},R_{2}]$ (where it has the largest amplitude). Thus, the spiral 
pattern appears nearly stationary. The growth rate $\gamma$ is also slow
relative to the local rotation, although the characteristic growth
time $\gamma^{-1} \simeq 10P_0$ is not very long.  
\begin{figure}
  \includegraphics[width=\linewidth]{figures/polarxy2_dens120_fargo}
  \caption{Cartesian visualisation of the $m=1$ surface density
    structure in the FARGO simulation initialised with only $m=1$
    perturbations. 
    \label{2d_fargo_viz}} 
\end{figure}

Next, we write $\Sigma_1 = 
|\Sigma_1|\exp{(\ii kR)}$, where $k$ is real, and assume the amplitude
$|\Sigma_1|$ varies slowly compared to the complex phase. This is the
main assumption in local theory. We measured $k$ numerically and plot
its normalised value in Fig. \ref{fargo_wavenumber}. We find 
\begin{align*}
  kR \sim \frac{\pi G \Sigma}{c_s^2}R \sim \frac{1}{hQ}, 
\end{align*}
where we used $Q\sim c_s\Omega/\pi G \Sigma$ and $R\Omega/c_s\sim
h^{-1}$. Since $Q=O(1)$ and $h\ll 1$ imply $|kR|\gg 1$, we 
can apply results from local theory
self-consistently. Fig. \ref{2d_fargo_viz} shows the $m=1$ spiral is
trailing, which is consistent with $k>0$.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_kr120_fargo}
  \caption{Normalised radial wavenumber of the $m=1$ spiral in 
    Fig. \ref{2d_fargo_viz}.\label{fargo_wavenumber}} 
\end{figure}

@@

Using the estimated value of $R_c$, we plot in
Fig. \ref{fargo_qbarrier} the quantity $\nu^2 - 1 + Q^{-2}$, which is
required to be positive in local theory for purely wave-like
solutions to the dispersion relation (Eq. \ref{dispersion}) when the
mode frequency is given. %Note that, with the above measured wavenumber $k$, 
%Eq. \ref{dispersion} requires $\nu^2 - 1 + Q^{-2}\sim0$ and the lower
%sign (long waves) to be taken. 
Fig. \ref{fargo_qbarrier} shows two $Q$-barriers located in
the inner disc, at $R_{Qb}=R_0$ and $R_{Qb}=1.6R_0$; the bounded region is indeed the
where the $m=1$ spiral develops. This suggests that the one-armed 
spiral is trapped. Note in this region, $\nu^2 - 1 + 
Q^{-2}\simeq 0.1\ll 1$, which is necessary for consistency with 
the measured wavenumber $k$ and Eq. \ref{wavenumber}.  
%Eq. \ref{dispersion} with the lower sign taken (i.e. long waves).  
There is one outer Lindblad resonance at $R_L\simeq
7.2R_0$. Thus, acoustic waves may be launched in $R\gtrsim 7.2R_0$ by
the spiral disturbance in the inner disc \citep{lin11b}. 
%implications: effect on dynamics in outer disc

@@

\subsection{Angular momentum exchange with the background disc}  
The wavenumber $k = \pi G\Sigma/c_s^2$ minimises the shifted frequency
$\sbar$ in the dispersion relation, Eq. \ref{dispersion}.  
%\emph{axisymmetric} perturbations. 
Physically, then, it is not
surprising to find perturbations of this wavenumber in a
self-gravitating disc where $Q\sim 1$. However, according the
local dispersion relation, $m=1$ perturbations are
formally stable for real $k$.

In order to track down the origin for the (slow) growth of the
$m=1$ spiral, we invoke angular momentum conservation for linear
perturbations, Eq. \ref{lin_ang_mom_cons}. Assuming fluxes are
negligible at the disc boundaries, we can integrate
Eq. \ref{lin_ang_mom_cons} to give
\begin{align}\label{baroclinic_torque_int}
  \frac{d}{dt}\underbrace{\int_{R_\mathrm{min}}^{R_\mathrm{max}}\jlin
    2\pi R dR}_{\jlintot} 
  =\int_{R\mathrm{min}}^{R\mathrm{max}}T_\mathrm{BG} 2\pi R dR, 
\end{align}
where we recall $T_\mathrm{BG}$ is the torque density due to the imposed
sound-speed profile (Eq. \ref{baroclinic_torque}). We explicitly 
compute both sides of Eq. \ref{baroclinic_torque_int} using
simulation data, and compare them in Fig. \ref{fargo_angmom_ex}. There
is a good match between the two torques, especially at early times
$t\lesssim110P_0$. The average discrepancy is $\simeq 5\%$. 
The match is less good later on, when the spiral
amplitude is no longer small ($\Delta\Sigma_1\sim 0.2$ at $t=110P_0$
and $\Delta\Sigma_1\sim 0.4$ by $t=120P_0$) and linear theory becomes
less valid. Fig. \ref{fargo_angmom_ex} confirms that the $m=1$ spiral
wave experiences a negative torque that further reduces its angular
momentum. This is consistent with angular momentum component
measurements (Fig. \ref {2d_angmom}).

@@

\subsection{Dependence on the imposed temperature profile} 
We now hypothesize the growth of the $m=1$ spiral mode is
associated with the imposed temperature gradient. The torque
responsible for its growth, Eq. \ref{baroclinic_torque}, is
proportional to $q$, the power-law index used to set the sound-speed
profile (Eq. \ref{sound-speed}). To test this, we carried out
simulations with $q\in[0,1]$ using only an $m=1$ initial 
perturbation. However, to keep the Toomre $Q$ profiles similar, we adjust
the surface density power-law index such that $s = (3+q)/2$.

Fig. \ref{fargo_varq} compares the $m=1$ spiral amplitudes as a
function of $q$. We indeed observe slower growth with decreasing
$q$. Although the figure indicates growth for the strictly isothermal
disc ($q=0$), we did not observe a coherent one-armed spiral upon
inspection of the $m=1$ surface density field. The growth in this case
may be associated with higher $m$ modes, which dominated the
simulation.    
% show result of isothermal calc?

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_fargo_varq}   
  \caption{Evolution of the $m=1$ spiral amplitude as a function of
    the imposed sound-speed gradient $q$. The maximum value of the
    $m=1$ surface density in $R\in[R_1,R_2]$ is shown. 
    \label{fargo_varq}} 
\end{figure}

We plot growth rates of the $m=1$ spiral as a function of $q$ in 
Fig. \ref{fargo_varq_growth}. The correlation can be fitted with a
linear relation
\begin{align*}
  \gamma \simeq \left[0.015 q - 7.9\times10^{-4}\right] \Omega_k(R_0). 
\end{align*}
% difficult to measure very small growth rates
This confirms that the temperature gradient is responsible for the
development of the one-armed spirals observed in our simulations. 
\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ratemax_fargo_varq}    
  \caption{Growth rates of the $m=1$ spiral mode as a function of the
    imposed sound-speed gradient $q$ (asterisks). A linear fit is also
    plotted (dotted line). 
    \label{fargo_varq_growth}} 
\end{figure}
%maybe extend the low q simulations to get better rates?

@@

Fig. \ref{fargo_varh_growth} shows the growth rates of the $m=1$
spiral in $R\in[R_1,R_2]$ as a function of $h$. Growth rates increases
with $h$, roughly as  
\begin{align}
  \gamma \simeq \left[0.10h + 8.3\times10^{-3}\right]\Omega_k(R_0).  
\end{align}
However, a linear fit is less good than for variable $q$ cases above. This
may be due to the change in the total disc mass when $h$ changes. We
find no qualitative difference between the spiral pattern that
emerges.

>>>>

\subsection{Reference run}
To obtain a picture of the overall disc evolution, we describe a 
fiducial run initialised with $M=10$ in Eq. \ref{randpert}. 
Fig. \ref{fargo_modeamp} plots evolution of the maximum 
non-axisymmetric surface density amplitudes in $R\in[R_{1},R_{2}]$
for $m\in[1,10]$. Snapshots from the simulation are shown in
Fig. \ref{fargo_2d}. 
At early times $t\lesssim100P_0$ the disc 
dominated by low-amplitude high-$m$ perturbations. The $m\geq4$ modes
growth initially and saturate (or decays) after $t=40P_0$. Notice the
low $m\leq 2$ modes decay initially, but grows between $t\in[20,40]P_0$,
possibly due to non-linear interaction of the high-$m$ modes  
\citep{laughlin96,laughlin97}. However, the $m=1$ mode begins to grow
again after $t=70P_0$, and eventually dominates the annulus.

@@

\subsection{Properties of the $m=1$ spiral and its growth}\label{fargo_m1}
In order to characterise the $m=1$ spiral that eventually
dominates, here we examine a simulation initialised with 
$M=1$ in Eq. \ref{randpert}. The one-armed spiral that emerges is more
coherent than the above simulation, which facilitates the analysis
below.

Fig. \ref{2d_fargo_viz} shows a snapshot of the $m=1$ surface
density. By measuring the $m=1$ surface density amplitude and its
pattern speed, we obtain a co-rotation radius and growth rate 
\begin{align*}
  &R_c \simeq 4.4R_0,\\
  &\gamma\simeq 0.014\Omega_k(R_0) = 0.13\Omega_p. 
\end{align*}
This one-armed spiral can be considered as low frequency because its 
pattern speed $\Omega_p \simeq 0.1\Omega_k(R_0)\lesssim 0.3\Omega$ in
$R\in[R_{1},R_{2}]$ (where it has the largest amplitude). Thus, the spiral 
pattern appears nearly stationary. The growth rate $\gamma$ is also slow
relative to the local rotation, although the characteristic growth
time $\gamma^{-1} \simeq 10P_0$ is not very long.  
\begin{figure}
  \includegraphics[width=\linewidth]{figures/polarxy2_dens120_fargo}
  \caption{Cartesian visualisation of the $m=1$ surface density
    structure in the FARGO simulation initialised with only $m=1$
    perturbations. 
    \label{2d_fargo_viz}} 
\end{figure}

Next, we write $\Sigma_1 = 
|\Sigma_1|\exp{(\ii kR)}$, where $k$ is real, and assume the amplitude
$|\Sigma_1|$ varies slowly compared to the complex phase. This is the
main assumption in local theory. We calculate $k$ numerically and plot
its normalised value in Fig. \ref{fargo_wavenumber}. We find 
\begin{align*}
  kR \sim \frac{\pi G \Sigma}{c_s^2}R \sim \frac{1}{hQ}, 
\end{align*}
where we used $Q\sim c_s\Omega/\pi G \Sigma$ and $R\Omega/c_s\sim
h^{-1}$. Since $Q=O(1)$ and $h\ll 1$ imply $|kR|\gg 1$, we 
can apply results from local theory
self-consistently. Fig. \ref{2d_fargo_viz} shows the $m=1$ spiral is
trailing, which is consistent with $k>0$.

%snapshot shows it's tightly wound
\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_kr120_fargo}
  \caption{Normalised radial wavenumber of the $m=1$ spiral in 
    Fig. \ref{2d_fargo_viz}.\label{fargo_wavenumber}} 
\end{figure}

@@

Using the estimated value of $R_c$, we plot in
Fig. \ref{fargo_qbarrier} the quantity $\nu^2 - 1 + Q^{-2}$, which is
required to be positive in local theory for purely wave-like
solutions to the dispersion relation (Eq. \ref{dispersion}) when the
mode frequency is given. %Note that, with the above measured wavenumber $k$, 
%Eq. \ref{dispersion} requires $\nu^2 - 1 + Q^{-2}\sim0$ and the lower
%sign (long waves) to be taken. 
Fig. \ref{fargo_qbarrier} shows two $Q$-barriers located in
the inner disc, at $R_{Qb}=R_0$ and $R_{Qb}=1.6R_0$; the bounded region is indeed 
where the $m=1$ spiral develops. This suggests that the one-armed 
spiral is trapped. Note in this region, $\nu^2 - 1 + 
Q^{-2}\simeq 0.1\ll 1$, which is necessary for consistency with 
the measured wavenumber $k$ and Eq. \ref{wavenumber}.  
%Eq. \ref{dispersion} with the lower sign taken (i.e. long waves).  
There is one outer Lindblad resonance at $R_L\simeq
7.2R_0$. Thus, acoustic waves may be launched in $R\gtrsim 7.2R_0$ by
the spiral disturbance in the inner disc \citep{lin11b}. 
%implications: effect on dynamics in outer disc

@@

\subsubsection{Angular momentum exchange with the background disc}  
The wavenumber $k = \pi G\Sigma/c_s^2$ minimises the shifted frequency
$\sbar$ in the dispersion relation, Eq. \ref{dispersion}.  
Physically, then, it is not surprising to 
find perturbations of this wavenumber in a
self-gravitating disc where $Q\sim 1$. However, according the
local dispersion relation, $m=1$ perturbations are
formally stable for real $k$ and $Q>1$.

%most of the perturbation is inside co-rotation 
%so resonant interaction unlikely for instability

In order to track down the origin for the (slow) growth of the
$m=1$ spiral, we invoke angular momentum conservation for linear
perturbations, Eq. \ref{lin_ang_mom_cons}. Assuming fluxes are
negligible at the disc boundaries, we can integrate
Eq. \ref{lin_ang_mom_cons} to give
\begin{align}\label{baroclinic_torque_int}
  \frac{d}{dt}\underbrace{\int_{R_\mathrm{min}}^{R_\mathrm{max}}\jlin
    2\pi R dR}_{\jlintot} 
  =\int_{R\mathrm{min}}^{R\mathrm{max}}T_\mathrm{BG} 2\pi R dR, 
\end{align}
where we recall $T_\mathrm{BG}$ is the torque density due to the imposed
sound-speed profile (Eq. \ref{baroclinic_torque}). We explicitly 
compute both sides of Eq. \ref{baroclinic_torque_int} using
simulation data, and compare them in Fig. \ref{fargo_angmom_ex}. There
is a good match between the two torques, especially at early times
$t\lesssim110P_0$. The average discrepancy is $\simeq 5\%$. 
The match is less good later on, when the spiral
amplitude is no longer small ($\Delta\Sigma_1\sim 0.2$ at $t=110P_0$
and $\Delta\Sigma_1\sim 0.4$ by $t=120P_0$) and linear theory becomes
less applicable. Fig. \ref{fargo_angmom_ex} confirms that the $m=1$ spiral
wave experiences a negative torque that further reduces its angular
momentum. This is consistent with angular momentum component
measurements (Fig. \ref {2d_angmom}).

@@

\subsection{Dependence on the imposed temperature profile}
The simulations above suggest the growth of the $m=1$ spiral mode
can be attributed to the background torque applied on linear pertubrations
as a result of an imposed temperature gradient, as discussed in
\S\ref{global_cons}. According to Eq. \ref{baroclinic_torque}, this
torque is proportional to $q$, the power-law index used to set the
sound-speed profile (Eq. \ref{sound-speed}) in our simulations. 
We confirm this through a series of simulations with $q\in[0,1]$, 
initialised with $m=1$ perturbations. However, to keep the Toomre $Q$
profiles similar, we adjust the surface density power-law index such
that $s = (3+q)/2$.

Fig. \ref{fargo_varq} compares the $m=1$ spiral amplitudes as a
function of $q$. We indeed observe slower growth with decreasing
$q$. Although the figure indicates growth for the strictly isothermal
disc ($q=0$), we did not observe a coherent one-armed spiral upon
inspection of the $m=1$ surface density field. The growth in this case
may be associated with high-$m$ modes, which dominated the
simulation.

% difficult to measure  
% show result of isothermal calc?

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_fargo_varq}   
  \caption{Evolution of the $m=1$ spiral amplitude as a function of
    the imposed sound-speed gradient $q$. The maximum value of the
    $m=1$ surface density in $R\in[R_1,R_2]$ is shown. 
    \label{fargo_varq}} 
\end{figure}
We plot growth rates of the $m=1$ spiral as a function of $q$ in 
Fig. \ref{fargo_varq_growth}. The correlation can be fitted with a 
linear relation
\begin{align*}
  \gamma \simeq \left[0.015 q - 7.9\times10^{-4}\right] \Omega_k(R_0). 
\end{align*}
% difficult to measure very small growth rates
This shows that the temperature gradient is responsible for the
development of the one-armed spirals observed in our simulations.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ratemax_fargo_varq}    
  \caption{Growth rates of the $m=1$ spiral mode as a function of the
    imposed sound-speed gradient $q$ (asterisks). A linear fit is also 
    plotted (dotted line). 
    \label{fargo_varq_growth}} 
\end{figure}
%maybe extend the low q simulations to get better rates?

@@

Fig. \ref{fargo_varh_growth} shows the growth rates of the $m=1$
spiral in $R\in[R_1,R_2]$ as a function of $h$. Growth rates increases
with $h$, roughly as  
\begin{align*}
  \gamma \simeq \left[0.10h + 8.3\times10^{-3}\right]\Omega_k(R_0).  
\end{align*}
However, a linear fit is less good than for variable $q$ cases above. This
may be due to the change in the total disc mass when $h$ changes. We
find no qualitative difference between the spiral pattern that
emerges.

### 2014-11-26 paper/results_3d.tex
second draft

Visualisations of the 3D simulations are shown in   
Fig. \ref{3d_prelim} for the disc midplane and near the upper disc
boundary. The snapshots are chosen when the one-armed spirals in the two codes
have reached comparable amplitudes. The agreement between the codes,
as well as with the 2D simulations (Fig. \ref{2d_fargo_viz}), is
satisfactory. The largest spiral amplitude is found in the
self-gravitating region $R\in[R_1,R_2]$, independent of
height. However, notice the spiral pattern extends into 
the non-self-gravitating outer disc ($R>R_2$) at $z\sim 2H$, i.e.    
the disturbance becomes more global away from the midplane.

\begin{figure}
  \begin{center}
    \subfigure[ZEUS-MP]{
      \includegraphics[scale=0.305,clip=true,trim=0cm 0cm 0cm 0cm]{figures/polarxy2_dens015_z0}  
      \includegraphics[scale=0.305,clip=true,trim=0.8cm 0cm 0cm
      0cm]{figures/polarxy2_dens015_zmax}  
    }
    \subfigure[PLUTO]{
      \includegraphics[scale=0.305,clip=true,trim=0cm 0cm 0cm 0cm]{figures/pdiskxy_023_z0}
      \includegraphics[scale=0.305,clip=true,trim=0.8cm 0cm 0cm
      0cm]{figures/pdiskxy_023_zmax}
    }
  \end{center}
  \caption{Three-dimensional simulations using the ZEUS-MP (top) and 
    PLUTO (bottom) codes. The $m=1$ density component $\Delta\rho_1$
    at the midplane (left) and approximately 
    two scale-heights above the midplane (right) is shown. Here $\psi
    \equiv \pi/2 - \theta$ is the angular height  
    from the midplane. \label{3d_prelim}}   
\end{figure}

@@

Although Fig. \ref{3d_rz} appears to display significant vertical motion,
we measure the three-dimensionality parameter $\Theta < 10^{-2}$
(Eq. \ref{theta}), so vertical motion is insignificant compared to
horizontal motion. This supports a 2D approximation. On the other
hand, we find  $|v_z/c_s|\lesssim 0.2$ which, although sub-sonic, is
not very small.

>>>>

%more global than 2D?

Visualisations of the 3D simulations are shown in   
Fig. \ref{3d_prelim} for the disc midplane and near the upper disc
boundary. The snapshots are chosen when the one-armed spirals in the two codes
have reached comparable amplitudes. %The agreement between the codes,
%as well as with the 2D simulations (Fig. \ref{2d_fargo_viz}), is
%satisfactory. 
Both codes show similar one-armed patterns at either height, and the
midplane snapshot is similar to the 2D simulation
(Fig. \ref{fargo_2d}). 
The largest spiral amplitude is found in the
self-gravitating region $R\in[R_1,R_2]$, independent of
height. However, notice the spiral pattern extends into 
the non-self-gravitating outer disc ($R>R_2$) at $z\sim 2H$, i.e.    
the disturbance becomes more global away from the midplane.  
%2d only adequate for self-gravitaing region 
\begin{figure}
  \begin{center}
    \subfigure[ZEUS-MP]{
      \includegraphics[scale=0.305,clip=true,trim=0cm 0cm 0cm 0cm]{figures/polarxy2_dens015_z0}  
      \includegraphics[scale=0.305,clip=true,trim=0.8cm 0cm 0cm
      0cm]{figures/polarxy2_dens015_zmax}  
    }
    \subfigure[PLUTO]{
      \includegraphics[scale=0.305,clip=true,trim=0cm 0cm 0cm 0cm]{figures/pdiskxy_023_z0}
      \includegraphics[scale=0.305,clip=true,trim=0.8cm 0cm 0cm
      0cm]{figures/pdiskxy_023_zmax}
    }
  \end{center}
  \caption{Three-dimensional simulations using the ZEUS-MP (top) and 
    PLUTO (bottom) codes. The $m=1$ density component $\Delta\rho_1$
    at the midplane (left) and approximately 
    two scale-heights above the midplane (right) is shown. Here $\psi
    \equiv \pi/2 - \theta$ is the angular height  
    from the midplane. \label{3d_prelim}}   
\end{figure}

@@

Although Fig. \ref{3d_rz} appears to display significant vertical motion,
we measure the three-dimensionality parameter $\Theta < 10^{-2}$
(Eq. \ref{theta}), so vertical motions are insignificant compared to
horizontal motions. This supports a 2D approximation. On the other
hand, we find  $\mathrm{max}|v_z/c_s|\sim 0.2$ which, although
sub-sonic, is not very small.  
%may increase with height
